---
title: 'queryMed package: how to display drug interactions'
author: "Y. Rivault, O.Dameron, and N. Le Meur"
date: "version 0.1 - 17/04/2018"
output:
  html_document:
    df_print: paged
  pdf_document:
    citation_package: natbib
bibliography: queryMed.bib
vignette: |
  %VignetteIndexEntry{Vignette Title} 
  %VignetteEngine{knitr::rmarkdown} 
  usepackage[utf8]{inputenc}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Context

In the area of massive open-data access and big data in general, pharmaco-epidemiology and public health sciences are in need for computational tools. Researchers can now query large medical information systems such as medico-administrative databases and claim databases. Although those information systems are often well structured, their contents are highly codified with several medical terminologies and ontologies, which might be difficult to analyze by non-expert. Pharmaco-epidemiologists and public health scientists thus need batch translation of diverse medical codes like diagnostic codes (based on ICD9 or ICD10), medical procedures or drugs codes (based on ATC nomenclature). In addition to the translation of those codes, their annotation can help making the most of medico-administrative and claim databases analysis by public health researchers. For instance, in a patient care trajectory it might help identifying critical drug interactions that might impair the patient safety. It might also be used to predict evitable hospitalization due to inappropriated care trajectory.

Using ontologies in R has proven to be efficient and useful[@kurbatova11], notably in *omic fields of research. For example, many genomes (including the Human genome) are available for downlaod through the BioConductor repository. Those genomes are annotated and statistical analyses of enrichment of standardised terms (commun or closely related) within part of living organisms have help discovering new or impaired functions.

Simmilarly, knowledges in pharmacoepidemiology, drug interactions, indications and contraindications have been combined for clinical data analyzes [@pathak13]. However, while works like the Drug Indication Database (DID) and Drug Interaction Knowledge Base (DIKB) have pooled different sources of medical knowledge from the Web of Data [@ayvaz15; @sharp17], the use and merging of knowledge is still non trivial to perform due to the multitude of medical classifications and sources of knowledge.

Today the Linked Open Data and the Semantic Web provide technical solutions for the integration of distributed data, their interrogation and their interpretation. For instance, the Ressource Description Framework (RDF) standardise the representation of data and knowledge, and thus allows their sharing and reuse. Furthermore, SPARQL, an other standard from the Semantic Web, provides a way to query these structured data. So these technologies allow reasoning through shared knowledge in the form of ontologies, classifications, and medical thesauri. Although, querying Linked Data in R with SPARQL is feasible, it requiers expertise in the SPARQL language and data schema. 

To facilitate this step, we propose a tool to integrate medical ontologies programmatically in the R environment. The queryMed package provides functions and algorithms to query the different sources of medical knowledge representations from the Web of data and to link them to the main medical classifications, for the enrichment and the analysis of medical data. The proposed functions are of two sort: for expert and non-expert. In this vignette, we illustrate the interest of our package to display drug interactions as graphs and explore ATC structure and information.


## Applications

To illustrate how to visualize drug interactions with queryMed we explore the drugs that are listed to interact with the platelet antiaggregant CLOPIDOGREL.

First we load queryMed and the DIKB dataset that lists drug interactions gathered from various knowledge databases.
```{r load library, eval=T}
library(queryMed)
```

```{r load data, eval=T}
data(DIKB)
head(DIKB, n=3)
```

We subset the dataset and select only the rows where the drug of interest is the *clopidogrel* (object).

```{r clopidogrel data, eval=T}
clopData <- DIKB[DIKB$label5.atc.object=="CLOPIDOGREL",]
dim(clopData)
```

Overall it corresponds to `r nrow(clopData)` unique interactions. The main source is:

```{r source data, eval=T}
 colSums(clopData[, c("Kegg","FrenchDB","NDFRT","WorldVista")]>=1, na.rm=T)
```

To visualize the interactions we propose to use network representation. Currently, one can look at one drug (object) and its precipitants as a graph. The circle layout positionned the object of interest at the center (see Figure below). The thickness of the edge represents the fact that some interactions are reported by more than one source. The graphs below display the interactions listed by at least 2 sources (nbinteractions=2). Some drugs belong to the same ATC family and are represented by the same vertex color. When considering more general classes of the ATC nomenclature, the interactions concern 11 and 15 classes, respectively (Fig1B, Fig1C). Again some classes are characterized by a high number of interactions (size of the vertex) and sources (thickness of the edge), notably the anti-thrombotic and gastro-esophageal family. In the context of pharmacovigilance, one could then make assumptions about the value of expanding the study of interactions between clopidogrel and the 26 known drugs to those between clopidogrel and the families they belong to. 
For more then one drug a radarplot is proposed with the border of the graph being the score of interaction as define by the combinasion of the different sources and their relative level of precaution (see help page for more details).


```{r igraph, eval=T}
par(mfrow=c(2,2), mar=c(2,2,2,2)) 
pddi_plot("CLOPIDOGREL", level=5, nbinteractions=2)
text(-1.5,1 ,"A", cex=2)
pddi_plot("CLOPIDOGREL", level=4 , nbinteractions=2)
text(-1.5,1 ,"B", cex=2)
pddi_plot("CLOPIDOGREL", level=3, nbinteractions=2)
text(-2,1 ,"C", cex=2)
pddi_plot(c("CLOPIDOGREL","PRASURGEL"), level=5, plot="radar", nbinteractions=2)
text(-2,1 ,"D", cex=2)
```


```{r sessionInfo, eval=T}
sessionInfo()
```




